Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-02 overleaf/compiler.tex
Fixed figure in Sec. 4.
Changes to Sec. 4.
Shortening sections. Still way too long...

\ana{Convention: use \texttt{shared, plain}, {\sf code}, $i$ and $I$ for counters outside of code.}

@@

\begin{figure*}
  \includegraphics[width=0.9\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework. \ana{Change figure since we are not doing div-and-conquer.}}
  \label{fig:compiler_framework}
\end{figure*}

@@

Source syntax is essentially standard IMP syntax as shown below:
%\begin{figure}[t]
{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 e & ::= e \; \mathit{op} \; e \mid e \mid \x \mid  \texttt{const} \mid \texttt{A[$e$]} & \mathit{expression} \\
 % s &::= t \;\f & \mathit{field} \\
 %\md &::= t\;{\m}(t \;{\this}, \; t\;\x) \; \{\; \ol{t\;\y}\;\s;\;\return\;\y \; \}
 %                 & \mathit{method} \\
  s & ::= s;s \mid  & \mathit{sequence} \\
  & \x = e \mid \texttt{A[$e$]} = e \mid & \mathit{assignment \; stmt} \\
  & \texttt{for $i$ in range($I$)}: s \mid & \mathit{for \; stmt} \\
  & \texttt{if $e$: $s$ else: $s$} & \mathit{if \; stmt} \\
%   t &::= q \; {\sf C} & \mathit{qualified \; type}\\
%   q &::= \high \mid \poly \mid \low & \mathit{FlowCFL \; qualifier}\\
%    t' &::= q' \; \C & \mathit{field \; qualified \; type}\\
%   q' &::= \tainted \mid \poly & \mathit{field \; qualifier}\\

@@

The type system is mostly standard, and in our experience, a sweet spot between readability and expressivity.
The \texttt{shared} qualifier denotes shared values, i.e., ones shared among the parties and computed upon
under secure computation protocols; the \texttt{plain} qualifier denotes plaintext values.
Shared lists denote shared elements, but plaintext length. Subtyping is
$\texttt{plain} <: \texttt{shared}$, meaning that we can convert a plaintext value into a shared one,
but not vice versa. Subtyping on qualified types is again as expected, it is covariant in the qualifier
and invariant in the type: $\langle q_1 \; \tau_1 \rangle <: \langle q_2 \; \tau_2 \rangle$
iff $q_1 <: q_2$ and  $\tau_1 = \tau_2$.

@@

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[$i,j,k$] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[$i,j+2$] = ...} is not allowed.
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{enumerate}
Importantly, the arguments of MUX are restricted to base types, as passing array arguments
is not valid MOTION MPC. \ana{Check this.} This causes a slight inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes
annotations at the level of IMP Source (as in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a)), the
annotations propagate through the transformations and are checked at the 
level of MPC Source. Recall that the programmer annotates only inputs

@@

Our compiler translates from Source to SSA as follows:

%\begin{enumerate}
%  \item
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

        Convert the restricted-syntax AST to a three-address control-flow graph.
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
  %\item
        \paragraph{SSA conversion}:
        Convert the 3-address CFG to SSA with Cytron's algorithm.
%\end{enumerate}

>>>>

%\ana{Convention: use \texttt{shared, plain}, {\sf code}, $i$ and $I$ for counters outside of code.}

@@

\begin{figure*}
  \includegraphics[width=0.9\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework.}
  \label{fig:compiler_framework}
\end{figure*}

@@

Source syntax is essentially standard IMP syntax but with for-loops:
%\begin{figure}[t]
{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 e & ::= e \; \mathit{op} \; e \mid \x \mid  \texttt{const} \mid \texttt{A[$e$]} & \mathit{expression} \\
 % s &::= t \;\f & \mathit{field} \\
 %\md &::= t\;{\m}(t \;{\this}, \; t\;\x) \; \{\; \ol{t\;\y}\;\s;\;\return\;\y \; \}
 %                 & \mathit{method} \\
  s & ::= s;s \mid  & \mathit{sequence} \\
  & \x = e \mid \texttt{A[$e$]} = e \mid & \mathit{assignment \; stmt} \\
  & \texttt{for $i$ in range($I$)}: s \mid & \mathit{for \; stmt} \\
  & \texttt{if $e$: $s$ else: $s$} & \mathit{if \; stmt} \\
%   t &::= q \; {\sf C} & \mathit{qualified \; type}\\
%   q &::= \high \mid \poly \mid \low & \mathit{FlowCFL \; qualifier}\\
%    t' &::= q' \; \C & \mathit{field \; qualified \; type}\\
%   q' &::= \tainted \mid \poly & \mathit{field \; qualifier}\\

@@

The type system is mostly standard, and in our experience, a sweet spot between readability and expressivity.
The \texttt{shared} qualifier denotes shared values, i.e., ones shared among the parties and computed upon
under secure computation protocols; the \texttt{plain} qualifier denotes plaintext values.
Subtyping is $\texttt{plain} <: \texttt{shared}$, meaning that we can convert a plaintext value into a shared one,
but not vice versa. Subtyping on qualified types is again as expected, it is covariant in the qualifier
and invariant in the type: $\langle q_1 \; \tau_1 \rangle <: \langle q_2 \; \tau_2 \rangle$
iff $q_1 <: q_2$ and  $\tau_1 = \tau_2$.

@@

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[$i,j,k$] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[$i,j+2$] = ...} is not allowed. (This is again a restriction we imposed for convenience in our current implementation; we plan to extend the compiler with arbitrary write access.)
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{enumerate}
The arguments of MUX are restricted to base types. 
(Again, this is just a restriction of our current implementation that will be lifted.)
This causes an inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes
annotations at the level of IMP Source (as in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a)), the
annotations propagate through the transformations; annotations are inferred and checked with a taint analysis at the 
level of MPC Source. Tthe programmer annotates only inputs.

@@

Our compiler translates from Source to SSA as outlined below. Full details are in the extended version~\cite{Anon_TR}.

%\begin{enumerate}
%  \item
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        %This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

        Convert the restricted-syntax AST to a three-address control-flow graph. The step processes for-loops, if-statements and assignments as restricted by the syntax.
\begin{comment}
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
\end{comment}        
  %\item
        \paragraph{SSA conversion:}
        Convert 3-address CFG to SSA with Cytron's algorithm.
%\end{enumerate}
